Algorithms for Estimating Trends in Global Temperature Volatility

### 2018-09-04 aaai-2018/khodadadi-mcdonald-aaai19.tex
update figures
update appendix

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D\beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
With the penalty matrix $D$, the estimated $\beta$ will be piecewise
linear. \citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

@@

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^T h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly. To impose
this assumption, we estimate $h_t$ by solving the penalized, negative
log-likelihood: 
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ Dh }_1
\label{eq:l1tf_var}
\end{equation}
where $D$ has the same structure as above.

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is an
$T\times S$ vector and $D$ is replaced by $D_{ST} \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively\footnote{$N_s$ is obtained by
  counting the number of unique constraints at each location and at
  all times.}. The exact form of this matrix in clarified in Appendix
A in the Supplement. Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | D_{ST}h |.
\label{eq:l1tf_var_st_mat}
\end{equation}

@@

For a spatial grid of size $S$ and $T$ time steps, $D_{ST}$ will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $ST$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $S=90\times 360\approx 32,000$ spatial locations and
$T=3650$ time points, and so $D_{ST}$ has
approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP
algorithm, we need to solve a linear system of equations which
depends on $D_{ST}^\top D_{ST}$ (see appendix A and B). Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that $D_{ST}$ is a highly structured, sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

@@

\begin{figure}[tb]
  \centering	
  \includegraphics[height=.16\textheight]{Figures/convergence}  
  \includegraphics[height=.16\textheight]{Figures/modelComp_MAE}  
  \caption{Top: Convergence speed of linearized
    and consensus ADMM. Bottom: MAE for four models: admm\_opt: the
    proposed model with optimal values of $\lambda_t$ and $\lambda_s$,
    admm\_temp: no spatial penalty, admm\_sp: no temporal
    penalty.\attn{Fix these}} \label{fig:true_fitted_var}
\end{figure}

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}
  \caption{Left: The residuals of the time-series of Indianapolis
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details.} 
  \label{fig:bloom_estimatedSD}
\end{figure}

@@

The top row of \autoref{fig:avg_change_estimatedSD} shows the
detrended data, the estimated standard deviation and the yearly
average of these estimates for two cities in the US: a small
midwestern city (left) and San Diego (right). The estimated SD
captures the periodic behavior in the variance of the time-series. In
addition, the number of linear segments changes adaptively in each
time window depending on how fast the variance is changing.  
\begin{figure}[tb]
  \centering
  \includegraphics[width=.9\linewidth]{Figures/avg_logVar.pdf}\\
  \includegraphics[width=.9\linewidth]{Figures/avg_change_logVar.pdf}
  \caption{The
    average of the estimated variance over the northern hemisphere (left) and the
    change in the variance from 1961 to 2011 (right).}
\label{fig:avg_chg}
\end{figure}

@@

\appendix
\section{PDIP for $\ell_1$ Trend Filtering}
\label{sec:app_l1tf_var}

In this appendix we provide more details on how to solve the optimization problem \ref{eq:l1tf_var} using PDIP. The objective function is convex but not differentiable. Therefore, to be able to use PDIP we first need to derive the dual of this problem. We note that this is a generalized LASSO problem \citep{TibshiraniTaylor2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \norm{ Dx }_1$ is:

@@

where $f^*(\cdot)$ is the Fenchel conjugate of $f$: $f^*(u)=\max_x u^\top x-f(x)$. It is simple to show that

@@

for $w>0$, where $\mu_1$ and $\mu_2$ are dual variables for the $\ell_\infty$ constraint. Let $A=[\nabla r_{dual}^\top , \nabla r_{cent}^\top]^\top$. The newton step takes the following form

@@

The matrix $D$ in these equations have the following form:  $D_{total}^t=[D_{t}^t|D_{s}^t]$. The matrix $D_t$ is the following block-diagonal matrix and corresponds to the temporal penalty: 
 \begin{equation}
 D_t=\begin{bmatrix}
 D &  & \\ 
  & \ddots & \\
  &  & D
 \end{bmatrix}
 \label{eq:d_t_matrix}
 \end{equation}

where $D$ was defined in Section 2 of the main text. The number of the diagonal blocks is equal to the grid size $n_r \times n_c$. Each row of the matrix $D_s$ corresponds to one spatial constraint in Equation \eqref{eq:l1tf_var_st} in the text. For example, the first $T$ rows correspond to $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to $|h_{11t}-h_{12t}|$, and so on.

@@

\noindent where $n_b$ is the number of local variables in each sub-cube in \autoref{fig:data_cube}, and for ease of notation we have dropped the subscript $i$ and superscript $m$. Let $f(x)=\sum_{j=1}^{n_b} (x_j + y_j^2e^{-x_j}) + (\rho/2) \lVert x-\tilde{z} + u \lVert_2^2$. As it was explained in \autoref{sec:l1tf_var}, the dual of this optimization problem is: $\min_\nu f^*(-D^\top\nu)$ with the constraints $|\nu_k| \le \Lambda_k$. So to use PDIP we first need to compute the conjugate function $f^*(\cdot)$. We have:

@@

Having computed the conjugate function and its gradient and Jacobian, now we can use a number of convex optimization software packages which have an implementation of PDIP to perform the x-update step inside the ADMM loop. We chose the python API of the \texttt{cvxopt} package ~\citep{andersen_cvxopt:_2013}.

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=\columnwidth]{Figures/bloom_estimatedSD}
  \caption{Left: The residuals of the time-series of Indianapolis
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details.} 
  \label{fig:bloom_estimatedSD}
\end{figure}

>>>>

\citet{KimKoh2009} proposed $\ell_1$-trend filtering as a
method for estimating a smooth, time-varying trend. It is formulated
as the optimization problem 
% : a time-series $y_t$, $t=1,...,T$ is observed and we believe that
% it consists of a smooth trend and a stochastic component and the
% goal is to estimate the trend. The estimated trend is desired to be
% smooth, but at the same time, we seek small estimated residuals
% (stochastic component). These two objectives cannot be achieved
% simultaneously and so a trade-off between them should be made. In
% $\ell_1$-trend filtering, this trade-off is formulated as the
% following convex optimization problem: 
$
\min_{\beta} \frac{1}{2} \sum_{t=1}^{T} (y_t-\beta_t)^2+\lambda
\sum_{t=1}^{T-2} \left|\beta_t-2\beta_{t+1}+\beta_{t+2}\right| 
$
or equivalently:
\begin{equation}
\min_{\beta} \frac{1}{2} \norm{ y-\beta }_2^2+\lambda \norm{ D_t \beta}_1
\label{eq:l1tf}
\end{equation}
 where $y=\{y_t\}_{t=1}^T$ is an observed time-series, $\beta\in \mathbb{R}^T$ is the smooth trend,
 $D_t$ is a $(T-2)\times T$ matrix, and $\lambda$ is a tuning parameter
 which balances fidelity to the data (small errors in the first term)
 with a desire for smoothness.  
% The \textit{penalty matrix} is:
% \begin{equation}
% D=\begin{bmatrix}
% 1 & -2 & 1 &  &  &  &\\ 
% & 1 & -2 & 1 &  &  &\\ 
% &  & \ddots & \ddots & \ddots  &  &\\ 
% &  & & 1 & -2 & 1 &  \\ 
% &  &  &   & 1 & -2 & 1 
% \end{bmatrix}
% \label{eq:d_matrix}
% \end{equation}
%The first term in \eqref{eq:l1tf} penalizes large residuals while the second term encourages smooth estimated $\beta$.
In Appendix A, we specify the form of the matrix $D_t$ which results in piecewise
linear estimated $\beta$. \citet{KimKoh2009} proposed a specialized primal-dual
interior point (PDIP) algorithm for solving \eqref{eq:l1tf}. From a
statistical perspective, \eqref{eq:l1tf} can be viewed as a constrained maximum
likelihood problem with independent observations from a normal
distribution with common variance, $y_t \sim \mbox{N}(\beta_t,
\sigma^2)$, subject to a piecewise linear constraint on
$\beta$. Alternatively, solutions to \eqref{eq:l1tf} are equivalent to
maximum a posteriori Bayesian estimators based on Gaussian likelihood
with a special Laplace prior distribution on $\beta$. Note that the
structure of the estimator is determined by the penalty function
$\lambda\norm{D_t\beta}_1$ rather than any parametric trend
assumptions---autoregressive, moving average, sinusoidal seasonal
component, etc. The resulting trend is therefore essentially
nonparametric in the same way that splines are. In fact,
using squared $\ell_2$ norm as the penalty instead of $\ell_1$ results exactly in
regression splines.

@@

Inspired by the $\ell_1$-trend filtering algorithm, we propose a
non-parametric model for estimating the variance of a time-series. To
this end, we assume that at each time step $t$, there is a parameter
$h_t$ such that the observations $y_t$
are independent normal variables with zero mean and variance
$\exp(h_t)$. The negative log-likelihood of the observed data in this
model is $l(y\given h) \propto -\sum_{t=1}^T h_t - y_t^2e^{-h_t}$. Crucially,
we assume that the parameters $h_t$ vary smoothly. To impose
this assumption, we estimate $h_t$ by solving the penalized, negative
log-likelihood: 
\begin{equation}
\min_h -l(y\given h)+\lambda \norm{ D_th }_1
\label{eq:l1tf_var}
\end{equation}
where $D_t$ has the same structure as above.

@@

The first term in the objective is proportional to the negative
log-likelihood, the second is the temporal penalty for the
time-series at each location $(i,j)$, while the third and fourth,
penalize the difference between the estimated variance of two
vertically and horizontally adjacent points, respectively. The spatial
component of this
penalty is a special case of trend filtering on
graphs~\cite{WangSharpnack2016} which penalizes the difference between
the estimated values of the signal on the connected nodes (though the
likelihood is different). As before,
we can write \eqref{eq:l1tf_var_st} in matrix form where $h$ is an
$T\times S$ vector and $D_t$ is replaced by $D \in
\mathbb{R}^{(N_t+N_s) \times (T \cdot S)	}$, where $N_t=S \cdot
(T-2)$ and $N_s=T \cdot (2n_rn_c-n_r)$ are the number of temporal and
spatial constraints, respectively\footnote{$N_s$ is obtained by
  counting the number of unique constraints at each location and at
  all times.}. The exact form of this matrix in clarified in Appendix
A in the Supplement. Then, as we have two different tuning parameters
for the temporal and spatial components, we write $\Lambda
=\left[\lambda_1\one_{N_t}^\top,\;
  \lambda_2\one_{N_s}^\top\right]^\top$ leading
to:\footnote{Throughout the paper, we use $|x|$ for both scalars and
  vectors. For vectors we use this to denote a vector obtained by
  taking the absolute value of each entry of $x$.}  
% let $h$ be a
% vector whose first $T$ entries are $h_{11t}$ for $t=1,...,T$, its next
% $T$ entries are $h_{21t}$ and so on, then the optimization problem in
% the matrix form is as
\begin{equation}
\min_h -l(y\given h)+ \Lambda^\top | Dh |.
\label{eq:l1tf_var_st_mat}
\end{equation}

@@

For a spatial grid of size $S$ and $T$ time steps, $D$ in equation \ref{eq:l1tf_var_st_mat} will have
$3Tn_rn_c-2n_rn_c-Tn_r$ rows and $S \cdot T$ columns. For a $1^\circ\times
1^\circ$ grid over the entire northern hemisphere and daily data over
10 years, we have $S=90\times 360\approx 32,000$ spatial locations and
$T=3650$ time points, and so $D$ has
approximately $10^8$ columns and $10^8$ rows. In each step of the PDIP
algorithm, we need to solve a linear system of equations which
depends on $D^\top D$ (see appendix A and B). Therefore,
applying the PDIP directly is infeasible for our data.\footnote{We
  note that $D$ is a highly structured, sparse matrix, but, unlike
  trend filtering alone, it is not banded. We are unaware of general
  linear algebra techniques for inverting such matrix, despite our
  best efforts.}

@@

\begin{figure}[tb]
  \centering	
  \includegraphics[height=.16\textheight]{Figures/convergence}  
  \includegraphics[height=.16\textheight]{Figures/modelComp_MAE}  
  \caption{Top: Convergence speed of linearized
    and consensus ADMM. Bottom: MAE for four models: our method: the
    proposed model with optimal values of $\lambda_t$ and $\lambda_s$,
    temporal: no spatial penalty, spatial: no temporal
    penalty.\attn{Fix these}} \label{fig:true_fitted_var}
\end{figure}

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=.8 \columnwidth]{Figures/indy_estimatedSD_1}
  \includegraphics[width=.45 \columnwidth]{Figures/indy_estimatedSD_2}
  \includegraphics[width=.45 \columnwidth]{Figures/indy_estimatedSD_3}  
  \caption{Left: The residuals of the time-series of Indianapolis
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details. \attn{Fix the caption (top-bottom or left right)}} 
  \label{fig:bloom_estimatedSD}
\end{figure}

@@

The top row of \autoref{fig:avg_change_estimatedSD} shows the
detrended data, the estimated standard deviation and the yearly
average of these estimates for two cities in the US: a small
midwestern city (left) and San Diego (right). The estimated SD
captures the periodic behavior in the variance of the time-series. In
addition, the number of linear segments changes adaptively in each
time window depending on how fast the variance is changing.  
\begin{figure}[tb]
  \centering
  \includegraphics[width=.9\linewidth]{Figures/avg_logVar.pdf}\\
  \includegraphics[width=.9\linewidth]{Figures/avg_change_logVar.pdf}
  \caption{The
    average of the detrended estimated variance over the northern hemisphere (top) and the
    change in the variance from 1961 to 2011 (bottom).}
\label{fig:avg_chg}
\end{figure}

@@

\appendix
\section{PDIP for $\ell_1$ Trend Filtering of variance}
\label{sec:app_l1tf_var}

In this appendix we provide more details on how to solve the optimization problem with the objective specified in equation \ref{eq:l1tf_var} using PDIP. The objective function is convex but not differentiable. Therefore, to be able to use PDIP we first need to derive the dual of this problem. We note that this is a generalized LASSO problem \citep{TibshiraniTaylor2011}. The dual of a generalized LASSO with the objective $f(x)+\lambda \norm{ Dx }_1$ is:

@@

\noindent where $f^*(\cdot)$ is the Fenchel conjugate of $f$: $f^*(u)=\max_x u^\top x-f(x)$. It is simple to show that for the objective function of Equation \ref{eq:l1tf_var}

@@

\noindent for $w>0$, where $\mu_1$ and $\mu_2$ are dual variables for the $\ell_\infty$ constraint. Let $A=[\nabla r_{dual}^\top , \nabla r_{cent}^\top]^\top$. The newton step takes the following form

@@

Having computed the conjugate function and its gradient and Jacobian, now we can use a number of convex optimization software packages which have an implementation of PDIP to solve the optimization problem with the objective function \ref{eq:l1tf_var}. We chose the python API of the \texttt{cvxopt} package ~\citep{andersen_cvxopt:_2013}.

@@

\noindent where $n_b$ is the number of local variables in each sub-cube in \autoref{fig:data_cube}, and for ease of notation we have dropped the subscript $i$ and superscript $m$.

The matrix $D$ has the following form:  $D^t=[D_{temp}^t|D_{spat}^t]$. The matrix $D_{temp}$ is the following block-diagonal matrix and corresponds to the temporal penalty: 
\begin{equation}
D_{temp}=\begin{bmatrix}
D_t &  & \\ 
& \ddots & \\
&  & D_t
\end{bmatrix}
\label{eq:d_t_matrix}
\end{equation}

\noindent where $D_t$ was first introduced in Section 2 of the main text and has the following form:

\begin{equation}
 D_t=\begin{bmatrix}
 1 & -2 & 1 &  &  &  &\\ 
 & 1 & -2 & 1 &  &  &\\ 
 &  & \ddots & \ddots & \ddots  &  &\\ 
 &  & & 1 & -2 & 1 &  \\ 
 &  &  &   & 1 & -2 & 1 
 \end{bmatrix}
\label{eq:d_matrix}
\end{equation}

The number of the diagonal blocks in $D_{temp}$ is equal to the grid size $n_r \times n_c$. Each row of the matrix $D_{spat}$ corresponds to one spatial constraint in Equation \eqref{eq:l1tf_var_st} in the text. For example, the first $T$ rows correspond to $|h_{11t}-h_{21t}|$ for $t=1,...,T$, the next $T$ rows correspond to $|h_{11t}-h_{12t}|$, and so on.

This optimization problem, is again a generalized LASSO problem with $f(x)=\sum_{j=1}^{n_b} (x_j + y_j^2e^{-x_j}) + (\rho/2) \lVert x-\tilde{z} + u \lVert_2^2$.

As it was explained in Appendix A, the dual of this optimization problem is: $\min_\nu f^*(-D^\top\nu)$ with the constraints $|\nu_k| \le \Lambda_k$. To use PDIP we first need to compute the conjugate function $f^*(\cdot)$. We have:

@@

@@

\begin{figure}[tb]
  \centering
  \includegraphics[width=\columnwidth]{Figures/indy_estimatedSD_1}
  \caption{Left: The residuals of the time-series of Indianapolis
    (averaged weekly) and the estimated SD obtained from the method of
    \autoref{sec:l1tf_var} (red). Middle: the estimated SDs (red) and
    their annual average (black) without imposing the long horizon
    penalty. Right: the same as middle panel but here the long horizon
    penalty is imposed. See the text for more details.} 
  \label{fig:bloom_estimatedSD}
\end{figure}
